---
title: "BIOST 536: Homework 1"
subtitle: "Department of Biostatistics @ University of Washington"
author: 
- Alejandro Hernandez
date: "October 5, 2024"
output: pdf_document
---

```{r setup, include=F}
# setup options
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
options(knitr.kable.NA = '-')
labs = knitr::all_labels()
labs = labs[!labs %in% c("setup", "llm_appendix", "allcode")]
```

```{r load-libraries-data, include=F}
# clear environment
rm(list=ls())

# load relevant packages
library(dplyr)      # data frame manipulation
library(knitr)      # table formatting
library(gtsummary)  # "table 1" summary
library(sjPlot)     # model coefficient table
# library(tidyverse)

# load data
aml <- read.csv("../data/leukemia_data.csv") %>% 
  # read in select variables as factors
  dplyr::mutate_at(vars(tx, sex, eval, cr, status, bmtx, incl), as.factor)
# create boolean variables
aml$tx2 <- ifelse(aml$tx=="D", 0, 1)
aml$sex2 <- ifelse(aml$sex=="M", 0, 1)
aml$cr2 <- ifelse(aml$cr=="N", 0, 1)
# rename levels of select factors
levels(aml$tx) <- c("Daunorubicin", "Idarubicin")
levels(aml$sex) <- c("Female", "Male")
levels(aml$cr) <- c("Did not complete remission", "Completed remission")
```

## Background

The following questions are related to the evaluation of a clinical trial in patients with acute myelogenous leukemia (AML). The primary endpoint of the clinical trial was induction of complete remission (binary outcome); the two treatments being compared were the newly synthesized anthracycline *idarubicin* and the standard anthracycline agent *daunorubicin*. This assignment ignores certain sequential aspects of the original study and analyzes the data as if investigators always intended to analyze 130 subjects. The complete data and documentation are on the course Canvas site (not publicly accessible).

## Descriptive Statistics

1. Provide suitable descriptive statistics for this dataset as might be presented in Table 1 of a manuscript appearing in the medical literature.

```{r question-1, fig.cap="Table 1"}

## ===============
## Question 1
## ===============
# create "Table 1" summary
aml %>%
  dplyr::select(tx, cr, sex, age, fab, karn, wbc, plt, hgb, status) %>%
  gtsummary::tbl_summary(by = tx, missing = "no") %>%
  bold_labels()
```


## Measures of Association

2. Summarize the data in a 2x2 table where outcome D is complete remission and exposure E is treatment group.  Estimate the RR, RD, and OR. Which of the three summary measures do you think AML patients would be most interested in?

```{r question-2}

## ===============
## Question 2
## ===============
# 2x2 table to summarize the number of patients who completed remission 
# in each treatment arm
table(aml$cr, aml$tx) %>%
  knitr::kable(caption = "Contingency table of patients who completed remission 
               in each treatment arm")

# prob of complete remission given unexposed (standard treatment)
pRgU <- mean(subset(aml, tx2==0)$cr2)
# prob of complete remission given exposed (new treatment)
pRgE <- mean(subset(aml, tx2==1)$cr2)

# relative risk of complete remission given exposure
rr.E <- pRgE / pRgU
# risk difference of complete remission given exposure
rd.E <- pRgE - pRgU
# odds ratio of complete remission given exposure
odds.E <- pRgE / (1 - pRgE)
odds.notE <- pRgU / (1 - pRgU)
or.E <- odds.E / odds.notE
```

The **relative risk** of complete remission given exposure is `r round(rr.E, 3)`. Then, according to our clinical trial, we estimate the likelihood of reaching complete remission to be approximately `r round(rr.E, 1)` times greater in the arm treated with idarubicin than in the arm treated with daunorubicin.
In simple terms: the proportion of patients in our sample who reached complete remission as a part of the new treatment group is approximately `r round(rr.E, 1)` times greater than the proportion of patients who reached complete remission as part of the standard treatment group.

The **risk difference** of complete remission given exposure is `r round(rd.E, 3)`. We estimate the probability of reaching complete remission to differ between our treatment groups by approximately `r round(100*rd.E)`%, with the idarubicin-treated arm having greater chances of completing remission. 

The **odds ratio** of complete remission given exposure is `r round(or.E, 3)`. We estimate the odds of reaching complete remission to be approximately `r round(or.E, 1)` times greater in the arm treated with idarubicin than in the arm treated with daunorubicin.

3. Summarize the data in a pair of 2x2 tables as done in Lecture 2, where D is complete remission, E is treatment group, and the covariate is sex.

```{r question-3}

## ===============
## Question 3
## ===============
# 2x2 table to summarize the number of patients who completed remission 
# in each treatment arm, by sex
subset(aml, sex == "Female", select = c(cr, tx)) %>%
  table() %>%
  kable(caption = "Contingency table of female patients who completed remission
        in each treatment arm")

subset(aml, sex == "Male", select = c(cr, tx)) %>%
  table() %>%
  kable(caption = "Contingency table of male patients who completed remission
        in each treatment arm")
```

## Logistic Regression

4. Perform a logistic regression analysis to assess the treatment effect of idarubicin compared to daunorubicin adjusted for sex.  In other words, estimate the sex-adjusted OR and present in language suitable for scientific publication.

```{r question-4, results=FALSE}

## ===============
## Question 4
## ===============
mod.lr <- glm(cr ~ tx + sex, family = binomial(), data=aml)
exp.coef <- exp(coef(mod.lr))
exp.confint <- exp(confint(mod.lr))
                
# print model coefficients with sjPlot (not displayed in PDF)
sjPlot::tab_model(mod.lr, title = "Logistic regression of complete remission",
                  show.r2 = F, show.aic = T)
```

We estimate that the odds of complete remission from acute myelogenous leukemia (AML) is `r round(exp.coef[2], 2)` times greater for a population treated with idarubicin compared to a population of the same sex treated with daunorubicin (95% CI for odds ratio: `r round(exp.confint[2,1], 2)` - `r round(exp.confint[2,2], 2)`).

This analysis of our sample suggests the newly synthesized anthracycline, idarubicin, has a statistically significant, positive effect on the induction of complete AML remission. However, our confidence interval indicates that observing a weak effect within a patient population would not be unusual. 

5. Using the subset of data on males, perform a logistic regression analysis to assess the treatment effect of idarubicin compared to daunorubicin for males. Repeat for females.

```{r question-5, results=FALSE}

## ===============
## Question 5
## ===============
mod.lr2 <- glm(cr ~ tx, family = binomial(), data=subset(aml, sex2==0))
exp.coef2 <- exp(coef(mod.lr2))
exp.confint2 <- exp(confint(mod.lr2))

mod.lr3 <- glm(cr ~ tx, family = binomial(), data=subset(aml, sex2==1))
exp.coef3 <- exp(coef(mod.lr3))
exp.confint3 <- exp(confint(mod.lr3))

# print model coefficients with sjPlot (not displayed in PDF)
tab_model(mod.lr2, title = "Logistic regression of complete remission among
          male patients", show.r2 = F, show.aic = T)
tab_model(mod.lr3, title = "Logistic regression of complete remission among 
          female patients", show.r2 = F, show.aic = T)
```

We estimate that the odds of complete AML remission is `r round(exp.coef2[2], 2)` times greater for a male population treated with idarubicin compared to a male population treated with daunorubicin (95% CI for odds ratio: `r round(exp.confint2[2,1], 2)` - `r round(exp.confint2[2,2], 2)`).

Comparing two treated female populations, we estimate the odds of complete AML remission to be `r round(exp.coef3[2], 2)` times greater for females receiving idarubicin than those receiving daunorubicin (95% CI for odds ratio: `r round(exp.confint3[2,1], 2)` - `r round(exp.confint3[2,2], 2)`).

Neither analyses of these subsamples suggest the idarubicin treatment has a statistically significant effect on the induction of complete remission of AML. Still, our confidence interval indicates that observing a strong effect within either group of patients would not be unusual.

6. You should have found that the sex-adjusted OR you obtained in Q4 is in between the two sex-specific OR you obtained in Q5.  Can you explain why this make sense?

The subgroups that fashioned the two conditional logistic regression models form a complete partition of the sample that fit the marginal logistic regression model; because logistic regression models the proportion of a population, the estimated proportion of the overall population will be an aggregate of the estimated proportions of the subpopulations.

7.	Fit a logistic regression model with treatment arm, sex, and their interaction. Use the model to estimate the treatment effect in males, and compare to your result to 5(a). Use the model to estimate the treatment effect in females, and compare to your result in 5(b). Comment on the similarity or difference.  In general, when you are asked for a point estimate you should include a confidence interval; however, for this problem you are not required to provide confidence intervals.


```{r question-7, results=FALSE}

## ===============
## Question 7
## ===============
mod.lr4 <- glm(cr ~ tx*sex, family=binomial(), data=aml)
coef4 <- coef(mod.lr4)
confint4 <- confint(mod.lr4)

# print model coefficients with sjPlot (not displayed in PDF)
tab_model(mod.lr4, title = "Logistic regression of complete remission", 
          show.r2 = F, show.aic = T)
```

We estimate that the odds of complete AML remission is `r round(exp(coef4[2] + coef4[4]), 2)` times greater for a male population treated with idarubicin compared to a male population treated with daunorubicin (95% CI for odds ratio: `r round(exp(confint4[2,1] + confint4[4,1]), 2)` - `r round(exp(confint4[2,2] + confint4[4,2]), 2)`).

Comparing two treated female populations, we estimate the odds of complete AML remission to be `r round(exp(coef4[2]), 1)` times greater for females receiving idarubicin than those receiving daunorubicin (95% CI for odds ratio: `r round(exp(confint4[2,1]), 2)` - `r round(exp(confint4[2,2]), 2)`).

8.

(a) Write the population attributable risk (as given in Lecture 1) as a function of the rate of exposure P[E] and the relative risk of disease RR.

(b) Suppose smokers have 22 times the risk of dying from lung cancer as non-smokers.  Consider a population of 35% smokers. Estimate the PAR for smoking and lung cancer death (point estimate only). Write a sentence presenting and interpreting the PAR.

(c) Suppose smokers have 22 times the risk of dying from lung cancer as non-smokers.  Consider a population of 5% smokers. Estimate the PAR for smoking and lung cancer death (point estimate only). Write a sentence presenting and interpreting the PAR.

(d) Comment on the difference between the PAR in (b) and (c).

```{r question-8}

## ===============
## Question 8
## ===============

```

9. Consider the R script `sim_casecontrolsampling.R` discussed in the first day of class.  A statistic not considered is the risk difference RD.  Would you expect the RD computed on a case-control sample to estimate the RD in the population?  Why or why not?  You should be able to answer this question based on the principles already discussed.  If you want to, you can modify the my.summary function to include the RD and examine the results.  However, this is not required for the homework.

```{r question-9}

## ===============
## Question 9
## ===============

```

**End of report. Code appendix begins on the next page.**

\pagebreak

## Code Appendix

```{r allcode, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```

**End of document.**
