---
title: "Example Markdown"
output: pdf_document
date: "2024-09-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#loading dataset
hiv <- read.csv("../data/hiv.csv")

#loading packages to be used
library(ggplot2)
library(scales)
library(ggthemes)
library(tidyr)
library(dplyr)
```

**Question 1.**

Including plots in R markdown.


```{r plot1, echo=FALSE, fig.cap = "Figure 1: Boxplot of distance to results center in kilometers by whether the participant collected their result."}

#note that caption will not show up unless output is html

# results as a factor variable 
hiv$results2 <- if_else(hiv$results == 1,"Yes","No")

# plot distance from center versus whether a participant collected their result
hiv %>% 
  ggplot(aes(x = as.factor(results2),  
             y = distvct)) +
    # mirrored histogram
    geom_violin() +
    # visual of median, IQR
    geom_boxplot(alpha = 0.4) +
    scale_color_colorblind() +
    theme_bw() +
    labs(y = "Distance from Test Result Center (km)", 
         x = "Whether Participant Collected Result")
```
**Question 2.**

Including a table in R markdown.

```{r table1}

hiv <- hiv %>% mutate(
  results2 = if_else(results == 1,"Collected Result","Did Not Collect Result"),
  incentive2 = if_else(incentive == 1,"Incentive","No Incentive"))

tab <- table(hiv$results2, hiv$incentive2)

knitr::kable(tab, caption = "Contingency table of whether results were collected
             by whether the participant recieved an incentive.")
```

**Question 3.**

Here is an example of using R output in an answer.


```{r logitreg}

mod <- glm(data = hiv, results ~ incentive, family = binomial())

```

We estimate that someone who received an incentive has `r round(exp(coef(mod)[2]),2)` (95% CI: `r round(exp(confint(mod)[2,1]),2)` - `r round(exp(confint(mod)[2,2]),2)`) times the odds of collecting their HIV test results compared to someone who did not receive an incentive.

